Improving the Stability of Type Soundness Proofs in Dafny

### 2023-10-17 paper/paper.tex
bunch more writing

\section{Introduction}

\begin{itemize}
    \item Explain the problem statement and setup:
    \item \begin{itemize}
      \item you've got a langauge, deeply embedded in dafny.
      \item This means that you have some big-step semantics for the language, implemented by a definitional interpreter which returns either a value, or an error. (show generic)
      \item You also have a typechecker, written in dafny: a boolean function which checks if a program has some type.  (show generic)
            This ideally implements some particular type system (presented declaratively, with rules). This usually works where there's one rule for each syntactic form, adn so the function pattern matches on the syntax, and makes recursive calls to check or infer the types of the premises.
      \item You want to show a soundness theorem, which states that well-typed programs are safe: either the program successfully evaluates to a value of the correct type, or it encounters one of some allowable class of errors. (show example, with examples of safety -- errors or no errors)
    \end{itemize}

    \item How do you do the proof?
    \begin{itemize}
      \item Proofs of type soundness on paper are by induction on typing derivations
      \item In dafny, this looks like pattern matching on the term, and making recursive calls for the IH. (show this scaffold)
      \item These proofs grow linearly with the size of the language: more cases for more kinds of expressions.
    \end{itemize}

    \item It's important that this proof be \emph{stable}. 
    \begin{itemize}
      \item Unlike mechanized metatheory developments in tools backed by a type-theoretic proof checker, Dafny's SMT-backed verification
      can be \emph{unstable}: the result of verifying the type soundness theorem
      can change, seemingly at random.

      \item Proof instability is empirically known to correlate with both (a) proof resource usage, and (b) proof resource usage \emph{variance}.
      Proofs which use a lot of resoruces are likely to be unstable, as are proofs which may vary greatly in their resource usage between runs.
    \end{itemize}

    \item The problem:
    \begin{itemize}
      \item Unfortunately, traditional direct-style proofs of type soundness are very unstable in dafny, and do not scale to realistic-scale langauges.
      \item (Color on the whack-a-mole issue) 
    \end{itemize}

    \item Contributions: 
    \begin{itemize}
      \item Demonstrate a technique for making mechanized metatheory proofs more stable (Section~\ref{sec:sts}) by axiomatizing \jwc{word} the evaluator and type checker.
      \item Evaluate the technique by comparing empirical metrics of proof stability against other methods of structuring the proof (Section~\ref{sec:eval}).
    \end{itemize}

    \item In Section~\ref{sec:discussion}, we'll discuss details of how the technique might apply in different situations, as well as how it was developed and deployed to build the type soundness proof for the cedar language.

    \jwc{Mention that this is used in Cedar, and cite the open-source development}

    \jwc{Make sure to namecheck the blog post}

\end{itemize}

\section{Stable Type Soundness Proofs}
\label{sec:sts}

\jwc{Build into this better.}
For a running example, let's consider the first-order expression language we
use for our evaluation in Section~\ref{sec:eval}.  The language has
types for integers, booleans, and records, all given as the Dafny type
\texttt{Ty}.  For terms, we have variables (represented as strings, since the
langauge has no binders), addition, subtraction, division, and, or,
conditionals, and record expressions and projections. These are all represented
by a Dafny ADT \texttt{Term}, which has a variant for each syntactic form.
To demonstrate how the language works, and how our technique applies, we focus
on just the behavior of the addition operation.

@@

\subsection*{Finding Stability}
\jwc{Fix sectionname}

Unfortunately, as we'll see in Section~\ref{sec:eval}, this proof approach is
not stable. We realized this while attemping to scale this style
of proof up to a realistic language. With enough operations and language
features, the proof resource usage varies wildly between runs,
enough that the verification will fail at random. At its core, this
instability arises from all of the afformentioned unguided reasoning the solver has to do about the
code of the typechecker and evaluator,
to get from the premise of the lemma to the premises of the IH, and then
from the conclusions of the IH to the conclusion of the case.
A patchwork solution is to add guidance in the form of assertions around the
recursive calls, to spell out the intermediate steps more directly.
\jwc{Words, about why this doesn't work in the long run}

@@

\section{Evaluation}
\label{sec:eval}
In this section, we demonstrate empirically that our technique as described in Section~\ref{sec:sts}
does yield more stable type soundness proofs. As a quantative proxy for stability, we will measure
the resource usage of our proofs. If a proof is expensive, or its cost varies wildly
between runs, this can be an indicator for future proof instability \jwc{Aaron Tomb, personal communication}.
While resource usage is the most important indicator of future proof instability, we also
base our conclusions on how much verification duration --- the wall-clock time it takes for the solver
to prove the VCs --- varies between runs.

@@

\subsection*{Results}
The graph in Figure~\ref{fig:cost-results} shows the mean verification \emph{resource usage} (and standard deviation, with whiskers), over the 250 verifications of each soundness theorem,
while the graph in Figure~\ref{fig:dur-results} shows verification \emph{durations}, in miliseconds. The two graphs tell essentially the same story.
\jwc{Re-Run the experiments!}
The highest cost and highest variance proof in all cases is the soundness proof for the typechecker in \texttt{monadic.dfy}. It strikes an unforuntate balance of being complicated code --- heavy use of monadic bind and lots of helper functions --- with little guidance in the proof.
The typechecker in \texttt{direct.dfy} is slightly cheaper to verify than the one in \texttt{monadic.dfy}, but it still has enormous variance. The cost seems to be lower because of all of the inlined definitions,
meaning that the verifier must reason about fewer functions while proving soundness.
The proof for the typechecker in \texttt{bidirectional.dfy} is cheaper and has lower variance than either of the previous two. This may be because while the bidirectionality increases the complexity of the typechecking algorithm,
this complexity \emph{requires} further guidance to the verifier for it to accept the proof at all, thereby improving stability on the whole.
As expected, the two proofs using the first and second halves of our technique
are by far the best. The soundness theorem using the fully specified typechecker and safety property (\texttt{semantic-invert.dfy})
is the best of all, with negligable variance in resource usage between runs of the verifier.

@@

\section{Discussion}

\subsubsection*{Scaling It Up}
A version of this proof technique was developed for the purposes of mechanizing the proof of type soundness
of the Cedar Language \jwc{cite} \footnote{https://github.com/cedar-policy/cedar-spec/tree/main/cedar-dafny},
the language underlying the Amazon Verified Permissions offering \footnote{https://aws.amazon.com/verified-permissions/}.
The technique described in this abstract is currently used in the type soundness proof of the reference typechecker with respect to the reference interpreter.
In Cedar, programs express access control policies, and the evaluator outputs an authorization decision: permit or deny.
Cedar's type system is a great deal more complex than that of our toy langauge, and includes a number of advanced type system features like ocurrence typing \jwc{cite: The Design and Implementation of Typed Scheme} and
singleton types \cite{https://dl.acm.org/doi/10.5555/645867.670933}.

>>>>

\section{Introduction}
Type soundness proofs for strongly-typed programming languages are often carried
out in type-theory based proof assistants like Coq or Agda. However, recent
small-scale experiements \cite{mayer-blog-post} and large-scale mechanized
metatheory developments \cite{cedar-proofs}.  have shown that Dafny is capable
enough as a proof assistant to encode type soundness proofs. In these
developments, Dafny serves both as the host language in which the target
language can be implemented, and a program verifier in which the mechanized
proof can be completed.

The usual way that a target language semantics is implemented in Dafny
is by way of a definitional interpreter \cite{definitional-interpreters}, a function \texttt{eval} which takes an
an abstract syntax tree \texttt{e} which represents a term in the target language, and returns either an error,
or the value \texttt{v} the term evaluates to. In principle, this implements some abstract semantics defined
by inference rules on paper: a semantic judgment $\texttt{e} \downarrow \texttt{v}$ holds if and only if \texttt{eval(e) = Ok(v)}.

The other half of a language implementation is a typechecker: a boolean function \texttt{check} which again takes a term \texttt{e}
and a type \texttt{t}, and returns true or false if \texttt{e} has type \texttt{t}. Again, this implements some abstract typing
judgment $\vdash \texttt{e} : \texttt{t}$, which is defined by way of inference rules on paper.

The goal of a type soundness proof is then to demonstrate that well-typed
programs do not go wrong: if $\vdash \texttt{e} : \texttt{t}$, then $\texttt{e}
\downarrow \texttt{v}$ for some value \texttt{v} of type \texttt{t}. When proven
on paper, this proof proceeds by an induction on the typing derivation of
$\vdash \texttt{e} : \texttt{t}$. For each rule defining the typing judgment, we
prove that if the premises do not go wrong, then the conclusion also does not go
wrong. Meanwhile, a \emph{mechanized} proof of this fact in Dafny takes the form
of a lemma which requires \texttt{check(e,t)}, and ensures that \texttt{eval(e)
= Ok(v)} and that \texttt{v} has type \texttt{t}. The proof of this lemma
necessarily looks like a pattern-match on the term \texttt{e}, making recursive
calls on the subterms to establish the inductive hyoptheses. For syntax-directed type systems
of the sort we will consider in this extended abstract\footnote{Strictly speaking, this is not a restriction of our tehcnique. The only practical way to implement a non-syntax-directed type system is to build a syntax-directed version, prove the two versions equivalent, and then implement the syntax-directed one.},
the structure of these two kinds of proofs are the same: each case of the pattern match encodes the proof for the corresponding typing rule for the corresponding syntactic form.

As such, the size of a type soundness proof in Dafny grows linearly with the
size of the language; more language features and more expressions mean more
cases.  This can make it hard to ensure that the proof is \emph{stable}.  Unlike
mechanized metatheory developments in tools backed by a type-theoretic proof
checker, Dafny's SMT-backed verification can be \emph{unstable}: the result of
verifying the type soundness theorem can change, seemingly at random.
Unfortunately, followign the proof structure described above leads to very unstable proofs in Dafny. Anecdotally, we have
found it impossible to scale this kind of proof to anything approximating a realistic language without encountering enormous
proof instability barriers.

In this extended abstract, we propose a new proof structure for mechanized type soundness proofs in Dafny which ensures proof stability.
By giving partial specifications of the evaluator and type checker, and proving type soundness relative to those specifications, we
can eliminate much of the \jwc{words}
Our technique has been used to verify \jwc{word}

In Section~\ref{sec:sts}, we present the technique by walking through a traditional-style proof of type soundness in Dafny.
Then in Section~\ref{sec:eval}, we evaluate the technique comparing empirical metrics of proof stability against other methods of structuring type soundness proofs, and other ways of writing a typechecker.
Lastly, in Section~\ref{sec:discussion}, we discuss how our technique has been deployed in practice, speculate about how it can be applied to even more complex languages,
and discuss the limitations and drawbacks.

\section{Stable Type Soundness Proofs}
\label{sec:sts}
For a running example, let's consider the first-order expression language we
use for our evaluation in Section~\ref{sec:eval}.  The language has
types for integers, booleans, and records, all given as the Dafny type
\texttt{Ty}.  For terms, we have variables (represented as strings, since the
langauge has no binders), addition, subtraction, division, and, or,
conditionals, and record expressions and projections. These are all represented
by a Dafny ADT \texttt{Term}, which has a variant for each syntactic form.
To demonstrate how the language works, and how our technique applies, we focus
on just the behavior of the addition operation.

@@

\subsection*{Finding Stability}
Unfortunately, as we'll see in Section~\ref{sec:eval}, this proof approach is
not stable. We realized this while attemping to scale this style
of proof up to a realistic language. With enough operations and language
features, the proof resource usage varies wildly between runs,
enough that the verification will fail at random. At its core, this
instability arises from all of the afformentioned unguided reasoning the solver has to do about the
code of the typechecker and evaluator,
to get from the premise of the lemma to the premises of the IH, and then
from the conclusions of the IH to the conclusion of the case.
A patchwork solution is to add guidance in the form of assertions around the
recursive calls, to spell out the intermediate steps more directly.
\jwc{Words, about why this doesn't work in the long run}

@@

\section{Evaluation}
\label{sec:eval}
In this section, we demonstrate empirically that our technique as described in Section~\ref{sec:sts}
does yield more stable type soundness proofs. As a quantative proxy for stability, we will measure
the resource usage of our proofs. If a proof is expensive, or its cost varies wildly
between runs, this can be an indicator for future proof instability \cite{stabilize},\cite{tomb}
While resource usage is the most important indicator of future proof instability, we also
base our conclusions on how much verification duration --- the wall-clock time it takes for the solver
to prove the VCs --- varies between runs.

@@

\subsection*{Results}
The graph in Figure~\ref{fig:cost-results} shows the mean verification \emph{resource usage} (and standard deviation, with whiskers), over the 250 verifications of each soundness theorem,
while the graph in Figure~\ref{fig:dur-results} shows verification \emph{durations}, in miliseconds. The two graphs tell essentially the same story.
The highest cost and highest variance proof in all cases is the soundness proof for the typechecker in \texttt{monadic.dfy}. It strikes an unforuntate balance of being complicated code --- heavy use of monadic bind and lots of helper functions --- with little guidance in the proof.
The typechecker in \texttt{direct.dfy} is slightly cheaper to verify than the one in \texttt{monadic.dfy}, but it still has enormous variance. The cost seems to be lower because of all of the inlined definitions,
meaning that the verifier must reason about fewer functions while proving soundness.
The proof for the typechecker in \texttt{bidirectional.dfy} is cheaper and has lower variance than either of the previous two. This may be because while the bidirectionality increases the complexity of the typechecking algorithm,
this complexity \emph{requires} further guidance to the verifier for it to accept the proof at all, thereby improving stability on the whole.
As expected, the two proofs using the first and second halves of our technique
are by far the best. The soundness theorem using the fully specified typechecker and safety property (\texttt{semantic-invert.dfy})
is the best of all, with negligable variance in resource usage between runs of the verifier.

@@

\section{Discussion}
\label{sec:discussion}

\subsubsection*{Scaling It Up}
A version of this proof technique was developed for the purposes of mechanizing the proof of type soundness
of the Cedar Language \cite{cedar}
the language underlying the Amazon Verified Permissions offering \footnote{https://aws.amazon.com/verified-permissions/}.
The technique described in this abstract is currently used in the type soundness proof of the reference typechecker with respect to the reference interpreter.
In Cedar, programs express access control policies, and the evaluator outputs an authorization decision: permit or deny.
Cedar's type system is a great deal more complex than that of our toy langauge, and includes a number of advanced type system features like ocurrence typing \cite{typed-scheme} and
singleton types \cite{sing-types}.
